6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
More cleanup
Proofing

\documentclass[twocolumn,10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,hyperref}
\usepackage{subcaption}

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit-rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a bit-rate
after an attempted packet send, and \texttt{RemoveStaleResults()},
which removes all statistics older than 10 seconds.

SampleRate sends at the rate with the lowest average transmission
time.  Every ten packets, it sends out a sample packet at a random rate
that has a smaller predicted transmission time than the current
rate.  SampleRate updates its transmission time estimates every ten
seconds based on feedback from the card about which packets succeeded
and at which rate they were sent.  If the current best rate experiences
four successive failures, SampleRate stops using that rate.

@@

Minstrel utilizes the multi-rate retry chain (mrr), which is an array
that tells the card which rates to try before reporting a failure.  The
retry chain makes failures extremely unlikely.  Minstrel has a mrr of
[\emph{highest throughput, next highest throughput, best probability,
    lowest base rate}] for most packets.  In essence, Minstrel tells
the card to send at the highest throughput rates but to fall back to
reliable rates upon failure.  The throughput and success probability
are calculated for each rate every 100ms.  To keep an accurate
estimate of throughput and probability for each rate, Minstrel sends
sample frames ten percent of the time.  If the randomly chosen rate has
a higher lossless throughput than the current optimal rate, the mrr is
as follows: [\emph{random, highest throughput, best probability,
    lowest base rate}].  If the random rate is slower than the optimal
rate, it is moved further back in the retry chain: [\emph{best
    throughput, random, best probability, lowest base rate}].  This
ensures that Minstrel never samples rates worse than the current
optimal rate unless the optimal rate experiences a failure.  See table
~\ref{table:1} for an overview of the mrr's used in Minstrel.

Minstrel's implementation is based on SampleRate's, so it also has
\texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a
\texttt{UpdateStats()} method that runs every
100ms.  \texttt{UpdateStats()} is home to some of the key differences
between Minstrel and SampleRate.  Instead of making decisions based on
the average transmission time, Minstrel uses
throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation}
where $T$ is the throughput of some rate $r$, $p$ is the probability
of successfully transmitting at $r$, and $tx\_time$ is the lossless
transmission time at $r$.

The probability $p$ of a successful transmission is calculated from
statistics collected by the \texttt{ProcessFeedback()} method, and is
done using an exponential weighted moving average, or EWMA.  The EWMA
creates a weighted average that weighs recent data more heavily than
old data.  The weighting for old data decreases exponentially.  An EWMA
is beneficial because it ensures that a sudden degradation in link
quality will create a rapid response in the probabilities due to the
increased weight of recent data.  Minstrel will waste less time sending
packets at a rate that no longer works, as compared to a simple
probability calculation.

\begin{table*}[htb]
  \centering
    \begin{tabular}[htb]{l|l|l|l}
    \textbf{Try} & \textbf{Normal}               & \textbf{Random (slower than best)} & \textbf{Random (faster than best)} \\ \hline
    1   & Best throughput      & Best throughput           & Random                    \\
    2   & Next best throughput & Random                    & Best throughput           \\
    3   & Highest probability  & Highest probability       & Highest probability       \\
    4   & Lowest rate          & Lowest rate               & Lowest rate               \\
    \end{tabular}
    \caption{Multi-rate retry chains for Minstrel}
\label{table:1}
\end{table*}

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chip.  However, MadWifi is
deprecated, and only runs on now-ancient network cards.  In addition to
the unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users.  Both ath5k and ath9k, which
were created to replace MadWifi, did not port the old bit-rate
selection system from MadWifi and instead opted to use the Linux
kernel's rate selection framework, which only implements Minstrel.
Porting SampleRate over to the new drivers would not have been
trivial, as there are vast differences in the assumed interfaces.

At this point, we decided to take a different approach and use an
approach similar to that used in Sprout~\cite{sprout}.  We modified
the ath9k driver to try bit rates uniformly and report packet
successes and failures, then analyze this data in user-space with
re-implementations of SampleRate and Minstrel.  Not only did this
lessen the re-implementation cost (since the algorithms were
re-implemented in user-space in Python, not in kernel-mode in C), but
it also provided reproducibility and allowed us to investigate
modifications to SampleRate and to Minstrel.

@@

Traces were collected from an ath9k driver modified to sample
bit-rates uniformly (forgoing the usual bit rate selection algorithm)
and to report successes and failure from each packet.  Each bit-rate
was sampled uniformly in time -- thus, more packets were sent at
higher bit-rates, since those bit-rates otherwise deliver packets
faster.  For each packet sent, the time to send was recorded, and the
bit rate for the next packet was chosen to be the bit-rate that had
the smallest total transmissions time.  This guarantees approximately
uniform selection of bit-rates.  The wireless card used was an
Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.

@@

In Minstrel, the frequency of probe packets, the window size, and the
EWMA weighting are the adjustable parameters.  The 3.3.8 Linux kernel
uses a probe frequency of 10\%, a window size of 100ms, and a EWMA
weighting of 0.75.  We use the same values.  Additionally, Minstrel also
estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()}
method from the \texttt{util.c} file in \texttt{net/mac80211}.  We
implemented the same function in Python.

@@

Bit rate selection algorithms strive to have a throughput near or
greater than the optimal constant rate.  For this
\texttt{clear\_1.dat} trace, SampleRate clearly meets this goal,
achieving an average throughput more than 1 Mbps higher than the
optimal constant rate.  In noisier situations, though, Minstrel tends
to be closer to the optimal constant rate.

@@

\subsection{Availability}
All of our code and collected trace data are available on GitHub:

@@

This repository contains the modified ath9k driver used to collect the
traces, the traces we collected, as well as the Python simulation
framework.

@@

Very little analysis has been done of the performance of bit-rate
selection algorithms on 802.11n networks.  802.11n introduces many new
rates, as well as multiple-input/multiple output, or MIMO,
capabilities.  There is an 802.11n version of Minstrel, but there is no
implementation of SampleRate.

@@

Hari Balakrishnan suggested that we use a broadcast approach when
collecting our traces, instead of link layer acknowledgments.  This
would entail having multiple computers listening at the broadcast
address, and compiling successes and failures from that information.
This approach would free us from certain idiosyncrasies of reading
data from the driver, and lead to possibly more accurate traces.  We
were unable to implement this due to a lack of time and ability to
acquire compatible hardware.

Finally, we are considering submitting our improvements to the
Minstrel algorithm, namely the BEWMA, as a kernel patch.  Currently is
implemented in Python, so we would have to port the code into C.

@@

Minstrel and SampleRate perform similarly.  The results suggest that
SampleRate tends to do slightly better, except in very lossy links
such as the client being far awat from the AP or actively moving.
Minproved, our improved version of Minstrel, solidly outperforms both
Minstrel and SampleRate.

It seems like the abandonment of SampleRate in the Linux kernel may
have been a bit hasty, since it seems to perform similarly to Minstrel
and it is a much simpler algorithm and thus less bug prone.  We are
not sure if there are adjustments to SampleRate that will bring it's
performance near Minproved, though.

@@

{\footnotesize \bibliographystyle{acm}
\bibliography{paper}}

>>>>

\documentclass[twocolumn,10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,subcaption,hyperref}

@@

SampleRate, first introduced in \cite{samplerate}, is a bit rate
selection algorithm which maintains estimates of average transmission
time for each potential bit rate.To keep these estimates up to date,
SampleRate periodically sends a packet at a randomly-selected bit
rate, and uses the success or failure of this packet to update the
average transmission time of that bit-rate.  SampleRate also abandons
a bit rate after four successive failures at that rate.  All
information about packets is maintained with a 10-second sliding window.

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit-rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a
bit-rate after an attempted packet send, and
\texttt{RemoveStaleResults()}, which removes the effect of packets
older than 10 seconds.

@@

Minstrel, developed specifically for the Linux kernel, attempts to
improve upon SampleRate to make it better suited for noisy or
dynamic environments.  Unlike SampleRate, Minstrel tracks raw
probabilities of successful transmission for each bit rate, computed
based on an exponentially-weighted moving average of 100ms windows.

Minstrel makes use the multi-rate retry chain (MRR), an array of bit rates
and number of attempts to make at each bit rate, that tells the card
which rates to try before reporting a failure.  The retry chain makes
failures extremely unlikely and allows Minstrel to choose fallback
rates when a packet does not succeed.  Minstrel sets the MRR to first
try the rate with the highest throughput, then the next-highest
throughput, then greatest probability of success, and finally attempts
to send a packet at the lowest, base rate.  In essence, Minstrel tells
the card to send at the highest throughput rates but to fall back to
reliable rates upon failure.

To keep an accurate estimate of throughput and probability for each
rate, Minstrel sends sample frames ten percent of the time.  However,
Minstrel tries to avoid sampling at a rate slower than the current
best rate, if the current best rate is still successfully
transmitting.  If the randomly chosen rate has a higher lossless
throughput than the current optimal rate, the MRR first lists the
sample rate, then the highest-throughput rate, then the
best-probability rate, and finally the base rate.  On the other hand,
if the random rate is slower than the optimal rate, the sample rate is
placed lower in the retry chain: first, the best throughput rate is
tried, and only if it fails does the sample rate then get attempted
(followed, as usual, by the best probability rate and the base rate).
This ensures that Minstrel never samples rates worse than the current
optimal rate unless the optimal rate experiences a failure.  See
Table~\ref{table:mrr} for an overview of the retry chains used in
Minstrel.

Minstrel implementation is based on SampleRate, so it also has
\texttt{ApplyRate()} and \texttt{ProcessFeedback()} methods, as well
as a \texttt{UpdateStats()} method that runs every 100ms.
\texttt{UpdateStats()} is home to some of the key differences between
Minstrel and SampleRate.  Instead of making decisions based on the
average transmission time, Minstrel uses throughput, computed with $$T
= \frac{p}{tx\_time}$$ where $T$ is the throughput of some rate $r$,
$p$ is the probability of successfully transmitting at $r$, and
$tx\_time$ is the computed lossless transmission time at rate $r$.

The probability $p$ of a successful transmission is calculated from
statistics collected by the \texttt{ProcessFeedback()} method, and is
done using an exponential weighted moving average, or EWMA.  The EWMA
creates a weighted average that weighs recent data more heavily than
old data, with the weighting for old data decreasing exponentially.
An EWMA ensures that a sudden degradation in link quality will create
a rapid response in the probabilities, making the EWMA-using Minstrel
perform better in dynamic environments.  Minstrel will waste less time
sending packets at a rate that no longer works, as compared to a
simple sliding-window probability calculation.

\begin{table*}[htb]
  \centering
    \begin{tabular}[htb]{l|l|l|l}
    \textbf{Try} & \textbf{Normal} &
    \textbf{Random (slower than best)} &
    \textbf{Random (faster than best)} \\ \hline
    1 & Best throughput      & Best throughput     & Random              \\
    2 & Next best throughput & Random              & Best throughput     \\
    3 & Highest probability  & Highest probability & Highest probability \\
    4 & Lowest rate          & Lowest rate         & Lowest rate         \\
    \end{tabular}
    \caption{Multi-rate retry chains for Minstrel}
\label{table:mrr}
\end{table*}

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chip.  However, MadWifi is deprecated,
and only runs on now-ancient network cards.  In addition to the
unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users.  Both ath5k and
\texttt{ath9k}, which were created to replace MadWifi, did not port
the old bit-rate selection system from MadWifi and instead opted to
use the Linux kernel's rate selection framework, which only implements
Minstrel.  Porting SampleRate over to the new drivers would not have
been trivial, as there are vast differences in the assumed interfaces.

At this point, we decided to take a different approach and use an
approach similar to that used in Sprout~\cite{sprout}.  We modified
the \texttt{ath9k} driver to try bit rates uniformly and report packet
successes and failures, then analyze this data in user-space with
re-implementations of SampleRate and Minstrel.  Not only did this
lessen the re-implementation cost (since the algorithms were
re-implemented in user-space in Python, not in kernel-mode in C), but
it also provided reproducibility and allowed us to investigate
modifications to SampleRate and to Minstrel.

@@

Traces were collected from an \texttt{ath9k} driver modified to sample
bit-rates uniformly (forgoing the usual bit rate selection algorithm)
and to report successes and failure from each packet.  Each bit-rate
was sampled uniformly in time -- thus, more packets were sent at
higher bit-rates, since those bit-rates otherwise deliver packets
faster.  For each packet sent, the time to send was recorded, and the
bit rate for the next packet was chosen to be the bit-rate that had
the smallest total transmissions time.  This guarantees approximately
uniform selection of bit-rates.  The wireless card used was an
Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.

@@

In Minstrel, the frequency of probe packets, the window size, and the
EWMA weighting are the adjustable parameters.  The 3.3.8 Linux kernel
uses a probe frequency of 10\%, a window size of 100ms, and a EWMA
weighting of 0.75.  We use the same values.  Additionally, Minstrel
also estimates $tx\_time$ using the \verb|ieee80211_frame_duration()|
method from the \texttt{util.c} file in \texttt{net/mac80211}.  We
implemented the same function in Python.

@@

Bit rate selection algorithms strive to have a throughput near or
greater than the optimal constant rate.  For this \verb|clear_1.dat|
trace, SampleRate clearly meets this goal, achieving an average
throughput more than 1 Mbps higher than the optimal constant rate.  In
noisier situations, though, Minstrel tends to be closer to the optimal
constant rate.

@@

\section{Availability}

All of our code and collected trace data are available on GitHub:

@@

This repository contains the modified \texttt{ath9k} driver used to
collect the traces, the traces we collected, as well as the Python
simulation framework.

@@

Very little analysis has been done of the performance of bit-rate
selection algorithms on 802.11n networks.  802.11n introduces many new
rates, as well as multiple-input/multiple output (MIMO) capabilities.
There is an 802.11n version of Minstrel, but there is no similar
implementation of SampleRate.

@@

Hari Balakrishnan suggested that we use 802.11 broadcasts to collect
our traces, instead of link layer acknowledgments.  We would have
multiple computers listening at the broadcast address, and recording
all received packets.  If packets were sent over UDP, each computer's
record of which packets it received would allow us to compute the
success probability for each packet for each rate.  This approach
would free us from certain idiosyncrasies of reading data from the
driver, and lead to possibly more accurate traces, as well as the
ability to test multiple scenarios at once.  We were unable to
implement this due to a lack of time and inability to acquire
compatible hardware.

Finally, we are considering submitting our improvements to the
Minstrel algorithm, namely the balanced EWMA, as a kernel patch.
Currently our changes implemented in Python, so we would have to port
our improvements to C.

@@

Minstrel and SampleRate perform similarly.  The results suggest that
SampleRate tends to perform slightly better, except in very lossy
links such as the client being far awat from the AP or actively
moving.  Minproved, our improved version of Minstrel, solidly
outperforms both Minstrel and SampleRate.

It seems that the replacement of SampleRate in the Linux kernel may
have been a hasty, since it seems to perform similarly to Minstrel and
it is a much simpler algorithm and thus simpler to adapt to changing
wireless standards.  We are not sure if adjustments to SampleRate can
bring its performance to match that of Minproved, however.

@@

\bibliographystyle{acm}
\bibliography{paper}
